A Practical Guide for Generating Reproducible and Programmatic Neuroimaging Visualizations

### 2023-02-10 manuscript.Rmd
fig update

---
title: Moving Towards Reproducible and Programmatic Generation of Neuroimaging Visualizations
authors:
  - name: Sidhant Chopra
    thanks: Corresponding author
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
    email: sidhant.chopra@yale.com
  - name: Loïc Labache
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Elvisha Dhamala
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Edwina R Orchard
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Avram Holmes
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
keywords:
  - Neuroimaging visualization
  - Reproducibility 
  - Programmatic figures
  - Open Science
  - R
  - Python
  - Brain Visualization
abstract: Neuroimaging visualization forms the centerpiece of interpretation and communication of scientific results, as well as data quality control. Often, these images and figures are produced by manually changing settings on Graphical User Interfaces (GUI). There now exist many well-documented code-based brain visualization tools that allow users to programmatically generate publication-ready figures directly within R, Python and MATLAB environments. Compared to figures generated using GUIs, programmatic figures are more replicable, flexible, interactive, and integrated with the scientific process. Here we outline the advantages of using programmatic neuroimaging visualization tools, provide a list of available tools across programming environments and include examples of voxel, vertex, region-of-interest and edge-level visualizations.

output: 
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
    keep_tex: true
    latex_engine: lualatex
  bookdown::word_document2:
    toc: false
linestretch: 2
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{setspace}\onehalfspacing
---

@@

The visualization of neuroimaging data is one of the primary ways in which we evaluate data quality, interpret results, and communicate findings. These visualizations are commonly produced using graphical user interface-based (GUI) tools where individual images are opened and, within each instance, display settings are manually changed until the desired output is reached. In large part, the choice to use GUI-based software has been driven by a perception of convenience, flexibility, and accessibility. However, there now exist many code-based software packages which are well-documented and often do not require high-level knowledge of programming, making them more accessible to the neuroimaging community (Table 1). These tools are flexible and allow for the generation of reproducible, high-quality, and publication-ready brain visualizations in only a few lines of code (Figure 1), especially within the R, Python and MATLAB environments. Here we review the rational for the wide-spread adoption of code-generated visualizations by highlighting major advantages in replicability, flexibility, and integration over GUI based tools. We also providing didactic examples of visualizations and associated code as a point of reference.

# Replicability

@@

Visualization scripts should reflect a core feature of open science. Given that brain figures regularly form the centerpiece of interpretation within papers, conference presentations, or news reports, making sure they can be reliably regenerated is crucial for knowledge generation and dissemination. By writing and sharing code used to generate brain visualizations, a direct and traceable link is established between the underlying data and the corresponding scientific figure. While this code doesn't necessarily reflect the validity or accuracy of the scientific finding, it allows for reproducibility, instilling transparency and robustness, while demonstrating a desire to further scientific knowledge. Some even consider publishing figures which cannot be replicated as closer to advertising, rather than science (Steel, 2013).

While some GUI-based tools have historically offered command-line access to generate replicable visualizations, they can lack both the flexibility to easily generate publication ready figures and the benefits, such as iteration, provided by your preferred programming environment. Likewise, other GUI-based tools offer replicability in the form of automatically generated batch scripts or in-built terminals, which are often idiosyncratic and lack documentation to make them easily usable or reproducible by those not familiar with the specific software.

# Flexibility

Being able to exactly replicate your figures via code has marked advantages beyond open science practices. In particular, the ability to reprogram inputs (such as statistical maps) and settings (such as color schemes, thresholds, and visual orientations) can streamline your entire scientific workflow. Changing inputs and settings via code allows for the easy production of multiple figures, such as those resulting from multiple analyses which require similar visualizations. A simple for-loop or copying and pasting the code with altered input and/or settings-of-interest can be a powerful method for exploring visualization options or rapidly creating multi-panel figures. Likewise, an arduous request from a reviewer or collaborator to alter the image processing or analysis becomes less of a burden when the associated figures can be re-generated with a few lines of code, as opposed to re-pasting and re-illustrating them manually. Having a code-base with modifiable inputs can mean that the generation of visualizations requires less time, energy and effort than image and instance specific GUI-based generation. This also makes it easier to generate consistent figures across subsequent projects. Keep in mind that the gains of writing code for your figures are cumulative, and in addition to improving your programming, you start to build a code-base for figure generation that you can continue to reuse and share throughout your scientific career.

Precise controls via code over visualization settings, such as color schemes, legend placement and camera angles, can provide you with much greater flexibility over visualizations. Nonetheless, part of the appeal of GUI-based tools is that the presets for such settings can provide a useful starting point and reduce the decision burden on novice users. However, similar presets are often available in the form of default settings across most code-based packages, negating the need for the user to manually enter each and every choice required for creating an image. Most code-based tools also come with documentation, with R-packages on the *CRAN* or *Neuroconductor* (Muschelli et al., 2019) repositories strictly requiring detailed guidance. Recent packages have started to include detailed beginner-friendly documentation in GitHub repositories, or even entire papers (e.g., Pham, Muschelli, & Mejia, 2022; Mowinckel & Vidal-Piñeiro, 2020; Huntenburg et at., 2017; Schäfer & Ecker, 2020) which provide examples of figures that can be used as starting points or templates for new users. As the popularity of code-sharing for figure increases, there will be a cornucopia of templates that can be used as the basis for new figures.

While brain visualizations are often thought of as the end results of analyses, they also form a vital part of quality control for imaging data. Tools to automatically detect artefacts, de-noise the data and generate derivatives are becoming more robust, but we are not yet at the stage where visualizing the data during processing is no longer necessary. Nonetheless, when working with large datasets such as Human Connectome Project (Van Essen et al., 2013) or UK BioBank (Sudlow et al., 2015), it is unfeasible to use traditional GUI-based tools to visually examine the data. The time it takes to open a single file and achieve the desired visualization settings vastly compounds when working with large datasets. Knowing how to programmatically generate brain visualizations can allow you to iterate your visualization code over each image of a large datasets making checking the quality of each data processing step achievable. The visual outputs of each iteration can be complied into accessible documents which can be easily scrolled, with more advanced usage allowing for the creation of interactive HTML reports, similar to those created by standardized data processing tools like *fmriprep* (Esteban et al., 2019). Increasing capacity to conduct visual quality control on larger datasets will improve the identification of processing errors and result in more reliable and valid findings.

# Integrative and Interactive Reporting

Often in neuroimaging studies, programming languages such as R, Python and MATLAB are used for stastical analysis and generating non-brain figures, but the brain figures are outsourced to separate GUI's based tools such as *FSLeyes*, *Freeview* or *ITK-snap*. Switching from your analysis environment to a GUI-based visualization process can be a cumbersome deviation from the scientific workflow. This can make debugging errors more difficult, as you have to regularly switch program to visually examine the results of any modifications or adjustments to prior analyses. Using the brain visualization tools that already exist within your chosen programming environment can provide instant visual feedback on the impact of modifications to processing or analysis.

Increasingly popular software such as *R Markdown*, *Quarto* and *Jupyter Notebook* allow for the mixing of prose and code in a single script, resulting in fully reproducible and publication ready papers. By using code-based tools available within your preferred environment, brain visualizations can be directly integrated and embedded within a paper or report. For instance, a fully reproducible version of the current paper can be found on [GitHub](https://github.com/sidchop/RepoNeuroVis). Some journals that publish neuroimaging studies are moving towards allowing the submission of reproducible manuscripts, including reproducible figures (e.g. *e-Life*, *Aperture Neuro*), with other journals like *F1000Research* and *GigaScience* even allowing on-demand re-running of code linked to the associated article using 'compute capsules' from the cloud-based platform Code Ocean (Code Ocean, 2021).

Neuroimaging data are often spatially 3D and can have multiple time points, adding a 4th dimension (e.g., functional imaging data). Thus, communicating findings or evaluating quality using static 2D slices is challenging, and may not be the best representation of the data, or the associted interpretations. While well-curated 3D renderings can help with spatial localisation (see Madan, 2015; Pernet & Madan, 2019), in the end, static images can only provide an incomplete representation of the data, and forces researchers to choose the "best" angle or slice to show, which often involves compromising one result to emphasize another. An added advantage of some of the code-based tools is that you can generate 'rich' media like interactive figures or animations, which allow users to zoom, rotate and scroll through slices. Interacting with a figure in this way can improve scientific communication of findings. Linking to or even embedding these videos or interactive figures in papers can greatly enhance the communication of findings and make your paper more engaging for the reader. Such rich brain visualizations lend themselves to being embedded or shared on science communication mediums beyond academic papers - such as presentations, websites and social media - all of which can promote the communication of your research with peers and reach larger audiences (Li and Xie, 2020). This last point is becoming increasingly salient as promoting science on social media has become a core medium for spreading discoveries, science communication to the public, and even a primary avenue for employment opportunities for early-career researchers (Baker, 2015; Lee, 2019). Overall, public engagement is one of the cornerstones of science, and the images we create are at the center of the process.

\
**Table 1. Examples of code-based neuroimaging visualizations tools that can be accessed directly within R, MATLAB and Python environments.**
|:-------------|:-----------|:-----------|:-----------|:-----------|:-----------|
|   [ANTsR](https://github.com/ANTsX/ANTsR)                                                               | ***+*** | ***+*** | ***+***       |         |             |
|   [brainconn](https://github.com/sidchop/brainconn)                                                     |         |         |               | ***+*** |             |
|   [brainR](https://github.com/muschellij2/brainR)                                                       | ***+*** |         | ***+***       |         |             |
|   [ciftitools](https://github.com/mandymejia/ciftiTools)                                                | ***+*** | ***+*** | ***+*****\*** |         |             |
|   [fsbrain](https://github.com/dfsp-spirit/fsbrain)                                                     | ***+*** | ***+*** | ***+*****\*** |         |             |
|   [ggseg](https://github.com/ggseg/ggseg)                                                               |         |         | ***+***       |         |             |
|   [neurobase](https://github.com/muschellij2/neurobase)                                                 | ***+*** |         |               |         |             |
|   [oro.nifti](https://github.com/muschellij2/oro.nifti)                                                 | ***+*** |         |               |         |             |
|   [ANTsPy](https://github.com/ANTsX/ANTsPy)                                                             | ***+*** | ***+*** | ***+***       |         |             |
|   [brainiak](https://github.com/brainiak/brainiak)                                                      | ***+*** |         |               |         |             |
|   [Brainplotlib](https://github.com/feilong/brainplotlib)                                               |         | ***+*** | ***+*****\*** |         |             |
|   [Brainspace](https://github.com/MICA-MNI/BrainSpace)/[surfplot](https://github.com/danjgale/surfplot) |         | ***+*** | ***+*****\*** |         |             |
|   [DIPY](https://github.com/dipy/dipy)                                                                  | ***+*** |         |               |         | ***+***     |
|   [ENIGMA TOOLBOX](https://github.com/MICA-MNI/ENIGMA)                                                  |         |         | ***+***       |         |             |
|   [FSLeyes](https://github.com/pauldmccarthy/fsleyes)                                                   | ***+*** | ***+*** | ***+***       |         | *+*         |
|   [ggseg](https://github.com/ggseg/python-ggseg)                                                        |         |         | ***+***       |         |             |
|   [graphpype](https://neuropycon.github.io/graphpype/)                                                  |         |         |               | ***+*** |             |
|   [MMVT](https://github.com/pelednoam/mmvt)                                                             |         | ***+*** | ***+***       | ***+*** |             |
|   [MNE](https://github.com/mne-tools/mne-python)                                                        | ***+*** | ***+*** | ***+***       |         |             |
|   [mrivis](https://github.com/raamana/mrivis)                                                           | ***+*** |         |               |         |             |
|   [NaNSlice](https://github.com/spinicist/nanslice)                                                     | ***+*** |         |               |         |             |
|   [netneurotools](https://github.com/netneurolab/netneurotools)                                         |         | ***+*** | ***+*****\*** |         |             |
|   [netplotbrain](https://github.com/wiheto/netplotbrain)                                                |         |         | ***+***       | ***+*** |             |
|   [nilearn](https://github.com/nilearn/nilearn)                                                         | ***+*** | ***+*** | ***+***       | ***+*** |             |
|   [niwidget](https://github.com/nipy/niwidgets)                                                         | ***+*** | ***+*** |               |         | ***+***     |
|   [Pycortex](https://github.com/gallantlab/pycortex)                                                    | ***+*** | ***+*** | ***+*****\*** |         |             |
|   [pySurfer](https://pysurfer.github.io/)                                                               |         | ***+*** | ***+*****\*** |         |             |
|   [surfice](https://github.com/neurolabusc/surf-ice)                                                    | ***+*** | ***+*** | ***+***       | ***+*** | ***+***     |
|   [Visbrain](https://github.com/EtienneCmb/visbrain)                                                    | ***+*** | ***+*** | ***+***       | ***+*** |             |
|   [BrainNetViewer](https://www.nitrc.org/projects/bnv/)                                                 | ***+*** |         | ***+***       | ***+*** |             |
|   [Brainspace](https://github.com/MICA-MNI/BrainSpace)                                                  |         | ***+*** | ***+*****\*** |         |             |
|   [Brainstorm](https://github.com/brainstorm-tools/brainstorm3)                                         | ***+*** | ***+*** | ***+***       |         |             |
|   [bspmview](https://www.bobspunt.com/software/bspmview/)                                               | ***+*** |         | ***+***       |         |             |
|   [CandlabCore](https://github.com/canlab/CanlabCore)                                                   | ***+*** |         | ***+***       |         |             |
|   [ECoG/fMRI Vis toolbox](https://github.com/Immiora/matlab-ecog-visualization)                         |         | ***+*** | ***+*****\*** |         |             |
|   [ENIGMA TOOLBOX](https://github.com/MICA-MNI/ENIGMA)                                                  |         |         | ***+***       |         |             |
|   [FieldTrip](https://www.fieldtriptoolbox.org/)                                                        | ***+*** | ***+*** |               |         |             |
|   [Lead-DBS](https://www.lead-dbs.org/)                                                                 | ***+*** |         | ***+***       |         |             |
|   [mni2fs](https://github.com/dprice80/mni2fs)                                                          |         | ***+*** |               |         |             |
|   [mrtools](https://github.com/justingardner/mrTools)                                                   | ***+*** | ***+*** | ***+*****\*** |         |             |
|   [plotSurfaceROIBoundary](https://github.com/StuartJO/plotSurfaceROIBoundary)                          |         | ***+*** | ***+*****\*** |         |             |
|   [Vistasoft](https://github.com/vistalab/vistasoft)                                                    | ***+*** | ***+*** | ***+***       |         | ***+***     |

Note: The tools listed contain functionality required to generate (at least close-to) publication-ready neuroimaging figures via user-entered code within R, MATLAB and Python environments. This list does not include cross-platform general purpose visualization software.\
***\**** Cortex only

# Examples of Neuroimaging Visualization Packages available in *R and Python*

The following two figures provide examples of voxel, vertex, ROI and edge-level figures generated within R (Figure 1) and Python (Figure 2) using open source, well documented and beginner-friendly packages. These are not an exhaustive representation of packages available for visualizing brain data in R and Python (See Table 1), rather, the figures aim to give the reader a sense of the available options, and an entry point to using them. All code used to compile the figures, as well as the contents of each panel are provided an accompanying [online repository.](https://github.com/sidchop/RepoNeuroVis)

```{=tex}
\newpage
\blandscape
```
```{r Figure1, echo=FALSE, out.height = '60%', out.width = '70%', fig.align='center'}

@@

***Figure 1. Examples of brain imaging visualizations made using R.*** A) Voxel-level statistical map thresholded and overlaid over a T1-weighted template image, with a single axial slice shown. Made using the the `ortho2` function from the `neurobase` package. B) A voxel-level cortical parcellation overlaid on a individual T1-weighted image, shown in 9-slice axial orientation. Made using the `overlay` function from the `neurobase` package. C) A CIFTI format surface ROI atlas with a corresponding statistic assigned to each region, with both hemispheres displayed on a inflated template surface in lateral view. Made using the `view_xifti_surface` from the `ciftiTools` package. D) A coronal cross-sectional rendering of subcortical structures where a value has been assigned to each region. Made using the `aseg` atlas from the `ggseg` package. E) A 3D rendering of 9 bilateral subcortical regions where a value has been assigned to each region. Made using the `aseg` atlas from the `ggseg3d` package. F) Medial and lateral views of a ROI atlas displayed on inflated cortical surface where a value has been assigned to each region. Made using the `glasser` atlas from the `ggsegGlasser` package, which was plotted using `ggseg`. G) Lateral view of a CIFTI format vertex-level data displayed on a inflated template surface. Made using the `view_xifti_surface` function from the `ciftiTools` package. H) Medial and lateral views of vertex-level data displayed on a individuals white matter surface. Made using the `vis.subject.morph.standard` function from the `fsbrain` package. I) A weighted and undirected graph plotted on top, left and front views of a schematic outline of a brain in MNI coordinate space. Made using the `brainconn` function from the `brainconn` package. All code used to compile this figure, as well as the contents of each panel are provided in an accompanying [online repository](https://github.com/sidchop/RepoNeuroVis).

@@

```{=tex}
\newpage
\blandscape
```
```{r Figure2, echo=FALSE, out.height = '60%', out.width = '80%', fig.align='center', dev='cairo_pdf'}
knitr::include_graphics("./data/python/Figure2.pdf")
```

>>>>

---
title: A Practical Guide for Generating Reproducible and Programmatic Neuroimaging Visualizations
authors:
  - name: Sidhant Chopra
    thanks: Corresponding author
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
    email: sidhant.chopra@yale.com
  - name: Loïc Labache
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Elvisha Dhamala
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Edwina R Orchard
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Avram Holmes
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
keywords:
  - Neuroimaging visualization
  - Reproducibility 
  - Programmatic figures
  - Open Science
  - R
  - Python
  - Brain Visualization
abstract: Neuroimaging visualizations form the centerpiece of the interpretation and communication of scientific results, and are also a cornerstone for data quality control. Often, these images and figures are produced by manually changing settings on Graphical User Interfaces (GUIs). There now exist many well-documented code-based brain visualization tools that allow users to programmatically generate publication-ready figures directly within R, Python and MATLAB environments. Here, we provide a rationale for the wide-spread adoption of code-generated brain visualizations by highlighting corresponding advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to generate these code-based brain visualizations. We also present a comprehensive table of tools currently available for programmatic brain visualizations and provide didactic examples of visualizations and associated code as a point of reference. Finally, we provide a web-app (https://sidchop.shinyapps.io/braincode/) which can generate simple code-templates as starting points for these visualizations.
output: 
  bookdown::word_document2:
    toc: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
    keep_tex: true
    latex_engine: lualatex
linestretch: 1.5
geometry: "left=1.5cm, right=1.5cm, top=1.5cm, bottom=1.5cm"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{setspace}\onehalfspacing
- \usepackage[left]{lineno}
- \modulolinenumbers[5]
- \linenumbers
---

@@

The visualization of neuroimaging data is one of the primary ways in which we evaluate data quality, interpret results, and communicate findings. These visualizations are commonly produced using graphical user interface-based (GUI) tools where individual images are opened and, within each instance, display settings are manually changed until the desired output is reached. In large part, the choice to use GUI-based software has been driven by a perception of convenience, flexibility, and accessibility. However, there now exist many code-based software packages which are well-documented and often do not require high-level knowledge of programming, making them more accessible to the neuroimaging community. These tools are flexible and allow for the generation of reproducible, high-quality, and publication-ready brain visualizations in only a few lines of code, especially within the R, Python and MATLAB environments. Here, we first discuss the rational for the wide-spread adoption of code-generated visualizations by highlighting major advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to make code-based brain visualizations and provide a web-app (<https://sidchop.shinyapps.io/braincode/>; Figure 4) which can generate simple code-templates as starting points for these visualizationds. We also present a comprehensive table of tools currently available for programmatic brain visualizations (Table 1) and provide didactic examples of visualizations and associated code as a point of reference (Figure 2-3). The focus of this guide is on human brain magnetic resonance imaging (MRI) data, but many of the principals discussed and tools provided will equally apply to visualizing data from other organs and imaging modalities such as EEG, MEG, PET and CT.

# Benefits of learning to generate code-based brain visualizations

## Replicability

@@

Code used for data visualization should reflect a core feature of open science. Given that brain figures regularly form the centerpiece of interpretation within papers, conference presentations, or news reports, making sure they can be reliably regenerated is crucial for knowledge generation and dissemination. By writing and sharing code used to generate brain visualizations, a direct and traceable link is established between the underlying data and the corresponding scientific figure. While this code doesn't necessarily reflect the validity or accuracy of the scientific finding, it allows for reproducibility, instilling transparency and robustness, while demonstrating a desire to further scientific knowledge. Some even consider publishing figures which cannot be replicated as closer to advertising, rather than science (Steel, 2013).

Notably, some GUI-based tools have historically offered command-line access to generate replicable visualizations (e.g., FreeView, FSLeyes, surfice), making their use potentially equally replicable to purely code-based tools. Use of these specialized command-line interfaces can provide a useful middle ground of those who have little experience with coding environments. Nonetheless, these interfaces often still have a learning curve, but can lack other advantages, such as iteration, provided by the preferred programming environment (see Sections \@ref(flexibility) and \@ref(integrative-and-interactive-reporting)). Likewise, other GUI-based tools offer replicability in the form of automatically generated batch scripts (text files containing lines of specialized commands that can be re-executed) or in-built terminals, which are often idiosyncratic and may lack documentation to make them easily usable or replicable by those not familiar with the specific software.

## Flexibility

Being able to exactly replicate a figures via code has marked advantages beyond open science practices. In particular, the ability to reprogram inputs (such as statistical maps) and settings (such as color schemes, thresholds, and visual orientations) can streamline the entire scientific workflow. Changing inputs and settings via code allows for the easy production of multiple figures, such as those resulting from multiple analyses which require similar visualizations. A simple for-loop or copying and pasting the code with altered input and/or settings-of-interest can be a powerful method for exploring visualization options or rapidly creating multi-panel figures. Likewise, an arduous request from a reviewer or collaborator to alter the image processing or analysis becomes less of a burden when the associated figures can be re-generated with a few lines of code, as opposed to re-pasting and re-illustrating them manually. Having a code-base with modifiable inputs can mean that the generation of visualizations requires less time, energy and effort than image and instance specific GUI-based generation. This also makes it easier to generate consistent figures across subsequent projects. Keep in mind that the gains of writing code for figures are cumulative, and in addition to improving programming skills, one can build a code-base for figure generation that can continue to be reused and shared throughout a scientific career.

Precise controls via code over visualization settings, such as color schemes, legend placement and camera angles, can provide much greater flexibility over visualizations. Nonetheless, part of the appeal of GUI-based tools is that the presets for such settings can provide a useful starting point and reduce the decision burden on novice users. However, similar presets are often available in the form of default settings across most code-based packages, negating the need for the user to manually enter each and every choice required for creating an image. Most code-based tools also come with documentation, with R-packages on the *CRAN* or *Neuroconductor* (Muschelli et al., 2019) repositories strictly requiring detailed guidance. Recent packages have started to include detailed beginner-friendly documentation in GitHub repositories, or even entire papers (e.g., Pham, Muschelli, & Mejia, 2022; Mowinckel & Vidal-Piñeiro, 2020; Huntenburg et at., 2017; Schäfer & Ecker, 2020) which provide examples of figures that can be used as starting points or templates for new users (also see Section\@ref(brain-code-a-web-app-for-generating-code-templates-for-brain-visualizations)). As the popularity of code-sharing for figure increases, there will be a cornucopia of templates that can be used as the basis for new figures.

While brain visualizations are often thought of as the end results of analyses, they also form a vital part of quality control for imaging data. Tools to automatically detect artefacts, de-noise the data and generate derivatives are becoming more robust, but we are not yet at the stage where visualizing the data during processing is no longer necessary. Nonetheless, when working with large datasets such as Human Connectome Project (Van Essen et al., 2013) or UK BioBank (Sudlow et al., 2015), it is simply not feasible to use traditional GUI-based tools to visually examine the data. The time it takes to open a single file and achieve the desired visualization settings vastly compounds when working with large datasets. Knowing how to programmatically generate brain visualizations can allow for iteration of visualization code over each image of a large datasets making checking the quality of each data processing step achievable. The visual outputs of each iteration can be complied into accessible documents which can be easily scrolled, with more advanced usage allowing for the creation of interactive HTML reports (see Section \@ref(generate-and-share-visualization)), similar to those created by standardized data processing tools like *fmriprep* (Esteban et al., 2019). Increasing capacity to conduct visual quality control on larger datasets will improve the identification of processing errors and result in more reliable and valid findings.

## Integrative and Interactive Reporting

Often in neuroimaging studies, programming languages such as R, Python and MATLAB are used for statistical analysis and generating non-brain figures, but the brain figures are outsourced to separate GUI-based tools such as *FSLeyes*, *Freeview* or *ITK-snap*. Increasingly popular software such as *R Markdown*, *Quarto* and *Jupyter Notebook* allow for the mixing of prose and code in a single script, resulting in fully reproducible and publication ready papers. By using code-based tools available within the preferred environment, brain visualizations can be directly integrated and embedded within a paper or report. For instance, a fully reproducible version of the current paper can be found on [GitHub](https://github.com/sidchop/RepoNeuroVis). Some journals that publish neuroimaging studies are moving towards allowing the submission of reproducible manuscripts, including reproducible figures (e.g. *e-Life*, *Aperture Neuro*), with other journals like *F1000Research* and *GigaScience* even allowing on-demand re-running of code linked to the associated article via cloud-based platforms like *Code Ocean* (Code Ocean, 2021).

Neuroimaging data are often spatially 3D and can have multiple time points, adding a 4th dimension (e.g., functional imaging data). Thus, communicating findings or evaluating quality using static 2D slices is challenging, and may not be the best representation of the data, or the associated interpretations. While well-curated 3D renderings can help with spatial localisation (see Madan, 2015; Pernet & Madan, 2019), in the end, static images can only provide an incomplete representation of the data, and they force researchers to choose the "best" angle or slice to show, which often involves compromising one result to emphasize another. An added advantage of some of the code-based tools is the generation of 'rich' media like interactive figures or animations, which allow users to zoom, rotate and scroll through slices. Interacting with a figure in this way can improve scientific communication of findings. Linking to or even embedding these videos or interactive figures in papers can greatly enhance the communication of findings and make papers more engaging for the reader. Such rich brain visualizations lend themselves to being embedded or shared on science communication mediums beyond academic papers - such as presentations, websites and social media - all of which can promote the communication of research with peers and reach larger audiences (Li and Xie, 2020). This last point is becoming increasingly salient as promoting science on social media has become a core medium for spreading discoveries via science communication to the public (Mueller-Herbst, et al., 2020; Smith & Seitz, 2019; Huber et al., 2019), to the research community (Luc et al., 2021; Quintana, 2020a), and even a primary avenue for employment opportunities for early-career researchers (Baker, 2015; Lee, 2019). Overall, public engagement is one of the cornerstones of science, and the images we create are at the center of the process.

# Generating Code-based Visualizations

In the following sections we outline the primary steps required when generating programmatic and reproducible human brain visualizations, and provide tools and heuristics to guide this process.

## Selecting your programming language

The first step (Figure 1) in generating code-based visualization is selecting the coding language you want to use. Three of the most popular languages are R, Python and MATLAB, all of which have many options for generating brain visualizations (see Table 1). This decision can be made based on what language you have prior experience with, or if you used one of these languages for the analysis part of your project, there can be advantages in using the same language for visualizations. Switching from your analysis environment to separate visualization environment, or a GUI-based visualization software, can be a cumbersome deviation from the scientific workflow. This can make debugging errors more difficult, as you must regularly switch program to visually examine the results of any modifications or adjustments to prior analyses. Using the brain visualization tools that already exist within your chosen programming environment can provide instant visual feedback on the impact of modifications to processing or analysis. Alternatively, the choice of programming language may be dictated by visualization and data-type you require. For instance, visualizing streamlines from tractography may not be currently available in the R environment (Table 1), and therefore requires the use of Python or MATLAB. Other constraints may include access to proprietary software like MATLAB, which would necessitate the use of open-source options such as R or Python.

```{r Figure1, echo=FALSE, out.height = '70%', out.width = '100%', fig.align='center'}
  knitr::include_graphics('images/Figure1.png')
```

**Figure 1** Primary steps to generating programmatic and reproducible human brain visualizations. Each step is outlined in Each step is outlined in the corresponding sub-section of Section \@ref(generating-code-based-visualizations). \newpage

## Identify your visualization type

Neuroimaging data and its derivatives can be visualized in multiple forms, which have different associated file-types and visualization requirements (see Section \@ref(identify-your-input-file-format.)). A brief description of more popular visualization types is provided below:

***Voxel*****.** A voxel is the 3D equivalent of a pixel. In neuroimaging, voxels are used to represent the intensity values of a 3D scan, such as an MRI or CT scan. The voxels can be rendered in different colors to indicate different tissue types or other features of interest. For example, in functional MRI (fMRI) scans, voxels can be colored based on their level of activation, to show which areas of the brain are more active during a specific task or for MRI, parts of the brain showing statistically significant anatomical alterations. These visualizations are often displayed as either slices in axial, sagittal or coronal planes (Fig2A-B; Fig3A-B; or a 3D rendering of the whole brain. Statistical values are displayed as overlays on template anatomical images, which follow a common stereotaxic coordinate system (e.g., MNI152), or on individual-specific anatomical images ('native-space').

***Vertex.*** A vertex is a point in 3D space that represents a location on the surface. In neuroimaging, vertices are used to create a mesh representation of a brain structure, such as the cerebral cortex or a subcortical structure. Each vertex has a set of coordinates that specify its location in 3D space and is connected to other vertices to form triangles, which make up the mesh. Vertices can be used to create a 3D visualization of the brain surface and color-coded based on different attributes such as sulcal depth, thickness, or functional activation. These visualizations are often displayed as 3D rendering of each hemisphere from medial and lateral views (Fig2C,G-H; Fig3C-D). Statistical values are displayed as overlays on template surfaces which follow a common stereotaxic coordinate system and fixed number of vertices (e.g., average; Fig2C,G; Fig3C-D), or on individual-specific surfaces that have been reconstructed using an anatomical image (Fig2H).

***Regions of Interest (ROI).*** In neuroimaging, ROIs are used to identify specific brain structures or areas that are relevant to the research question. ROIs can be defined using various methods, such as manual tracing, atlas-based parcellation, or functional activation patterns. Visualizing ROIs can be done by grouping and assigning the same statistical value or color to sets of voxels or vertices (Fig2B-C; Fig3B-C), or can be done through polygons. Polygonal brain visualizations are simple 2D or 3D shapes which graphically represent the brain or specific structures, but do not carry any additional information on spatial coordinates and only roughly estimate the shape of the brain and its structures. Each region of 2D (Fig 2D,F) and 3D (Fig 2E; Fig3E) polygon visualizations can be filled in with colors representing a region label or a statistical value.

***Edge.*** In the context of neuroimaging, an edge often represents a connection between two brain regions, which can be thought of as a line connecting two nodes (brain regions) in a network. To visualize edges, one common method is to use a matrix representation, often called a 'connectivity' or 'adjacency' matrix, which shows the strength of the connections between all pairs of brain regions. These matrices can be displayed as a heat map, where different colors represent different levels of connectivity. Another way to represent edges is using a 2D graph, where the vertices represent brain regions and the edges represent the connections between those regions, which displayed in a variety of ways, such as a 2D or 3D representation of the brain (Fig2I; Fig3F), with edges represented as lines connecting the vertices. The edges and nodes can also be color-coded based on properties such as the strength of the connection.

***Streamlines.*** Streamlines are a representation of the white matter fibers in the brain and are usually generated using tractography applied to diffusion MRI (dMRI). Streamlines are typically visualized as 3D lines that connect different points of the brain and can be used to create 3D visualizations of specific white matter tracts, or all streamlines between regions. They can be color-coded based on the direction of the fibers or properties of the tract, such as myelination. These visualizations can be overlaid on voxel or vertex level anatomical brain representations to provide an anatomical reference point.

The following two figures provide examples of voxel, vertex, ROI and edge-level visualizations generated within R (Figure 2) and Python (Figure 3) using open source, well documented and beginner-friendly packages. These are not an exhaustive representation of packages available for visualizing brain data in R and Python (see Table 1). Rather, the figures aim to give the reader a sense of the many options available, and an entry point to choosing the type of brain visualization needed (also see Section 4). All code used to compile the figures, as well as the contents of each panel are provided in the accompanying [online repository](https://github.com/sidchop/RepoNeuroVis).

```{=tex}
\newpage
\blandscape
```
```{r Figure2, echo=FALSE, out.height = '60%', out.width = '70%', fig.align='center'}

@@

***Figure 2. Examples of brain imaging visualizations made using R.*** A) Voxel-level statistical map thresholded and overlaid over a T1-weighted template image, with a single axial slice shown. Made using the the `ortho2` function from the `neurobase` package. B) A voxel-level cortical parcellation overlaid on a individual T1-weighted image, shown in 9-slice axial orientation. Made using the `overlay` function from the `neurobase` package. C) A CIFTI format surface ROI atlas with a corresponding statistic assigned to each region, with both hemispheres displayed on a inflated template surface in lateral view. Made using the `view_xifti_surface` from the `ciftiTools` package. D) A coronal cross-sectional rendering of subcortical structures where a value has been assigned to each region. Made using the `aseg` atlas from the `ggseg` package. E) A 3D rendering of 9 bilateral subcortical regions where a value has been assigned to each region. Made using the `aseg` atlas from the `ggseg3d` package. F) Medial and lateral views of a ROI atlas displayed on inflated cortical surface where a value has been assigned to each region. Made using the `glasser` atlas from the `ggsegGlasser` package, which was plotted using `ggseg`. G) Lateral view of a CIFTI format vertex-level data displayed on a inflated template surface. Made using the `view_xifti_surface` function from the `ciftiTools` package. H) Medial and lateral views of vertex-level data displayed on a individuals white matter surface. Made using the `vis.subject.morph.standard` function from the `fsbrain` package. I) A weighted and undirected graph plotted on top, left and front views of a schematic outline of a brain in MNI coordinate space. Made using the `brainconn` function from the `brainconn` package. All code used to compile this figure, as well as the contents of each panel are provided in an accompanying [online repository](https://github.com/sidchop/RepoNeuroVis).

@@

```{=tex}
\newpage
\blandscape
```
```{r Figure3, echo=FALSE, out.height = '70%', out.width = '90%', fig.align='center', dev='cairo_pdf'}
knitr::include_graphics("./data/python/Figure2.pdf")
```
